---
title: "Forecasting and Time Series Methods Notes 10"
author: "[Xiaorui Zhu](https://homepages.uc.edu/~zhuxr/) (zhuxr@mail.uc.edu)"
output: 
  html_document:
    code_folding: show
    fig_caption: yes
    highlight: tango
    number_sections: yes
    theme: readable
    toc: yes
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
body{ /* Normal  */
      font-size: 20px;
  }
code.r{ /* Code block */
    font-size: 16px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 16px;
}
h1 { /* Header 1 */
  color: DarkBlue;
}
h2 { /* Header 2 */
  color: DarkBlue;
}
h3 { /* Header 3 */
  color: DarkBlue;
}
</style>



```{r, warning=FALSE,message=FALSE, results = 'hide'}
# Install necessary packages
list_packages <- c("AER", "dynlm", "tidyverse", "fpp", "fpp2", 
                   "forecast", "readxl", "stargazer", "scales",
                   "quantmod", "urca", "vars", "tseries", "sarima")
new_packages <- list_packages[!(list_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load necessary packages
lapply(list_packages, require, character.only = TRUE)
```

# GARCH

# Neural Network Models

```{r}
fit <- nnetar(sunspotarea, lambda=0)
autoplot(forecast(fit,h=30))
sim <- ts(matrix(0, nrow=30L, ncol=9L),
  start=end(sunspotarea)[1L]+1L)
```

```{r, eval=FALSE}
for(i in seq(9))
  sim[,i] <- simulate(fit, nsim=30L)
autoplot(sunspotarea) + autolayer(sim)
fcast <- forecast(fit, PI=TRUE, h=30)
autoplot(fcast)
```

# ETS Models

Exponential Smoothing State Space Model

# STL Models

```{r, eval=FALSE}
p1 <- autoplot(calls) +
  ylab("Call volume") + xlab("Weeks") +
  scale_x_continuous(breaks=seq(1,33,by=2))
p2 <- autoplot(window(calls, end=4)) +
  ylab("Call volume") + xlab("Weeks") +
  scale_x_continuous(minor_breaks = seq(1,4,by=0.2))
gridExtra::grid.arrange(p1,p2)
calls %>% mstl() %>%
  autoplot() + xlab("Week")
calls %>%  stlf() %>%
  autoplot() + xlab("Week")
fit <- auto.arima(calls, seasonal=FALSE, lambda=0,
         xreg=fourier(calls, K=c(10,10)))
fit %>%
  forecast(xreg=fourier(calls, K=c(10,10), h=2*169)) %>%
  autoplot(include=5*169) +
    ylab("Call volume") + xlab("Weeks")
```


# TBATS Models
```{r, eval=FALSE}
data("calls")
calls %>%
  subset(start=length(calls)-2000) %>%
  tbats() -> fit2
fc2 <- forecast(fit2, h=2*169)
autoplot(fc2, include=5*169) +
  ylab("Call volume") + xlab("Weeks")
```

# Forecast Combinations
```{r}
data(auscafe)
train <- window(auscafe, end=c(2012,9))
h <- length(auscafe) - length(train)
ETS <- forecast(ets(train), h=h)
ARIMA <- forecast(auto.arima(train, lambda=0, biasadj=TRUE),
                  h=h)
STL <- stlf(train, lambda=0, h=h, biasadj=TRUE)
NNAR <- forecast(nnetar(train), h=h)
TBATS <- forecast(tbats(train, biasadj=TRUE), h=h)
Combination <- (ETS[["mean"]] + ARIMA[["mean"]] +
                  STL[["mean"]] + NNAR[["mean"]] + TBATS[["mean"]])/5
autoplot(auscafe) +
  autolayer(ETS, series="ETS", PI=FALSE) +
  autolayer(ARIMA, series="ARIMA", PI=FALSE) +
  autolayer(STL, series="STL", PI=FALSE) +
  autolayer(NNAR, series="NNAR", PI=FALSE) +
  autolayer(TBATS, series="TBATS", PI=FALSE) +
  autolayer(Combination, series="Combination") +
  xlab("Year") + ylab("$ billion") +
  ggtitle("Australian monthly expenditure on eating out")
c(ETS = accuracy(ETS, auscafe)["Test set","RMSE"],
  ARIMA = accuracy(ARIMA, auscafe)["Test set","RMSE"],
  `STL-ETS` = accuracy(STL, auscafe)["Test set","RMSE"],
  NNAR = accuracy(NNAR, auscafe)["Test set","RMSE"],
  TBATS = accuracy(TBATS, auscafe)["Test set","RMSE"],
  Combination =
    accuracy(Combination, auscafe)["Test set","RMSE"])
```

# Practicie Issues in Forecasting


## Time Series Constrained to an Interval


## Very Long and Very Short Time Series


